Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-27 paper.qmd
typo

@@

This is achieved by disallowing side effects and ensuring that builds operate
within hermetic (isolated) environments, with no hidden dependencies on global
system state. While this functional approach greatly enhances reproducibility,
it can sometimes introduce complexity for package maintainers, especially for
software that needs to download external assets during installation (e.g.,
certain Bioconductor packages). For end-users, however, this complexity is
largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
*variants* of a package on the same system.

[Nix]{.pkg} installs all packages into a centralised, content-addressed
directory called the [Nix]{.pkg} store (typically `/nix/store`). Each package
gets a unique subdirectory whose name includes a hash of all its inputs: source
code, dependencies, build instructions. This means
`/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can coexist
without conflicts.

@@

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang}, uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

@@

The complete, runnable source code for this example is available in the paper's
GitHub repository. A shell script, `run_polyglot_example.sh`, is provided to
automate the entire process from environment creation to final output, as
described at the end of this
section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/polyglot-example/run_polyglot_example.sh}].

>>>>

Despite these shortcomings, [Docker]{.pkg} remains valuable for specific use
cases: production deployment, CI/CD integration, and cross-platform
distribution. The two approaches are not mutually exclusive. [Nix]{.pkg} can be
installed and used inside [Docker]{.pkg}, combining [Nix]{.pkg}'s deterministic
builds with [Docker]{.pkg}'s deployment infrastructure. In this workflow,
[Nix]{.pkg} handles environment reproducibility during the build of the image,
while [Docker]{.pkg} serves as the distribution mechanism. This addresses
[Docker]{.pkg}'s build-time reproducibility issues while maintaining its deployment
advantages. However, for research workflows where interactive development is
primary and [Docker]{.pkg}'s complexity is a barrier, [Nix]{.pkg} alone may suffice.

Thankfully, it is entirely possible to combine both [Docker]{.pkg} and [Nix]{.pkg}:
[Nix]{.pkg} to setup reproducible environments and [Docker]{.pkg} for deployment
across the many different already existing infrastructures.

@@

This is achieved by disallowing side effects and ensuring that builds operate
within hermetic (isolated) environments, with no hidden dependencies on global
system state. The principles of functional package management (hermetic builds,
immutable artifacts, and declarative dependencies) are designed to produce
deterministic outputs. The effectiveness of this model was recently validated in
the most extensive study of build reproducibility to date. The effectiveness of
this model was recently validated at unprecedented scale by @malka2025, who
rebuilt 709,816 packages from historical nixpkgs snapshots (2017–2023). They
found rebuildability rates exceeding 99% and bitwise reproducibility rates
between 69% and 91%, with an upward trend over time. Importantly, they
identified that about 15% of unreproducible builds failed due to embedded build
timestamps, which a solvable issue rather than a fundamental limitation.

While this functional approach greatly enhances reproducibility, it can
sometimes introduce complexity for package maintainers, especially for software
that needs to download external assets during installation (e.g., certain
Bioconductor packages). For end-users, however, this complexity is largely
abstracted away thanks to both [rix]{.pkg} and [rixpress]{.pkg}.

Additionally, [Nix]{.pkg} supports multiple versions or *variants* of a package
on the same system. [Nix]{.pkg} installs all packages into a centralised,
content-addressed directory called the [Nix]{.pkg} store (typically
`/nix/store`). Each package gets a unique subdirectory whose name includes a
hash of all its inputs: source code, dependencies, build instructions. This
means `/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can
coexist without conflicts.

@@

To illustrate these capabilities, we present a polyglot pipeline that simulates
a Real Business Cycle (RBC) model in [Julia]{.proglang}, uses the resulting data
to train an [XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises
the results in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a
[Quarto]{.pkg} report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/case-study],
with additional details provided in the Appendix of this manuscript.

@@

## A Comparison of Pipeline Paradigms: Imperative vs. Declarative {#sec-comparison}

To illustrate the practical benefits of our framework, we implemented the RBC
pipeline example from Section @sec-rixpress using two distinct paradigms. The
first is a traditional, imperative stack combining Docker for environment
isolation, Make for workflow orchestration, and language-specific package
managers. The second is our proposed declarative framework using `rix` and
`rixpress`. While both approaches successfully execute the same polyglot
analysis to produce identical outputs, their comparison illuminates fundamental
differences in conceptual models, system complexity, and the nature of the
reproducibility guarantees they provide.

### The Imperative Composition of Specialized Tools

The traditional approach represents a composition of specialized tools, each
solving part of the reproducibility puzzle. This paradigm is entirely
imperative; the researcher must explicitly script the procedural steps for both
environment construction and workflow execution.

The codification of the environment is a procedural build process specified in a
`Dockerfile`. Successfully authoring such a file necessitates that the
researcher act as a system administrator, manually scripting a sequence of shell
commands to install each component. This includes managing `apt` repositories
for R, invoking external installers like `uv` and `rig` for Python and R
runtimes, and using `curl` and `tar` for Julia. Subsequently, separate commands
must be run to install packages within each language's ecosystem. The procedural
verbosity and intertwined concerns of this approach are reflected in the
resulting `Dockerfile`, which spans 108 lines to define the complete
environment.

Here is a sketch of this `Dockerfile`:

```dockerfile
# Add R repository and install specific version
RUN apt-get update && apt-get install -y software-properties-common
RUN add-apt-repository ppa:...
RUN curl -L https://rig.r-pkg.org/... | sh
RUN rig add 4.5.1

# Install Python with uv
RUN curl -LsSf https://astral.sh/uv/install.sh | sh
RUN uv python install 3.13

# Download and extract Julia
RUN curl -fsSL https://julialang-s3.julialang.org/... -o julia.tar.gz
RUN tar -xzf julia.tar.gz -C /opt/

# Install packages for each language separately
RUN echo 'options(repos = c(CRAN = ...))' > /root/.Rprofile
RUN Rscript -e 'install.packages(...)'

# Install Python packages using uv with specific versions for reproducibility.
RUN echo "pandas==2.3.3" > /tmp/requirements.txt && \
    echo "scikit-learn==1.7.2" >> /tmp/requirements.txt && \
    # ... more packages ...

RUN uv pip install --no-cache -r /tmp/requirements.txt && \
    rm /tmp/requirements.txt

# Install specific versions of Julia packages for reproducibility
RUN julia -e 'using Pkg; \
    Pkg.add(name="Arrow", version="2.8.0"); \
    # ... more packages ...'
```

This procedural philosophy extends from environment definition to workflow
orchestration. A `Makefile` defines the pipeline as a directed acyclic graph of
file dependencies. While robust, this model requires the researcher to manually
encode the dependency graph, tightly coupling the analytical logic to specific
file paths. Furthermore, since `make` is agnostic to the content of these files,
it provides no mechanism to enforce type or schema consistency between steps.

A direct consequence of this file-based orchestration model is the proliferation
of wrapper scripts. Each step in the `Makefile` invokes a script containing a
significant amount of procedural boilerplate unrelated to the scientific logic.
This includes parsing command-line arguments, handling data serialization and
deserialization, and managing file paths. The core analytical functions are thus
obscured by this necessary but distracting scaffolding.

The cumulative effect is a system of high structural complexity. The traditional
implementation requires nine separate files to fully specify the environment,
workflow, and analysis. This includes the `Dockerfile`, `Makefile`, three
wrapper scripts for orchestration, and three `functions` files containing the
core logic, in addition to the final report. The cognitive overhead is
substantial, as a user must comprehend the conventions and interactions of this
entire toolchain.

### A Unified Declarative Framework

The Nix-based framework inverts this imperative model. Instead of scripting the
procedural construction of the environment and workflow, the researcher declares
the desired end-state.

Environment specification is reduced to a single declarative statement in the
`gen-env.R` script. The researcher specifies *what* the environment should
contain, leaving the procedural implementation of *how* to build it entirely to
the Nix toolchain. The `gen-env.R` script replaces the 108-line `Dockerfile`,
abstracting away all system administration details. `rix` translates this
high-level specification into a formal Nix expression that deterministically
resolves all transitive dependencies for the complete polyglot environment.

```r
library(rix)
rix(
  date = "2025-10-14",
  r_pkgs = c("ggplot2", "dplyr", "arrow"),
  jl_conf = list(jl_version = "lts", ...),
  py_conf = list(py_version = "3.13", ...),
  ...
)
```

Workflow orchestration undergoes a similar conceptual shift from procedural to
declarative. The `gen-pipeline.R` script defines the pipeline not as a set of
file rules, but as a graph of data objects. Dependencies are not manually
specified but are automatically inferred by `rixpress` from the functional
relationships in the code. This object-centric model renders the procedural
wrapper scripts unnecessary. The scientific logic is expressed as a set of pure
functions that accept data structures as inputs and return them as outputs, free
of any boilerplate for file I/O or command-line parsing.

```python
# functions/functions.py
def train_and_predict_output(simulated_df: pd.DataFrame) -> pd.DataFrame:
    # ... scientific logic ...
    return format_results(predictions)
```

This declarative approach significantly reduces system complexity. The entire
implementation requires only six files: two for the high-level environment and
pipeline declarations, three containing the pure scientific functions, and the
final report. The reduction is not merely quantitative; the remaining files are
conceptually simpler, containing only analytical logic and specifications, with
all procedural scaffolding abstracted away by the framework.

### Reproducibility Guarantees: Space and Time

The critical distinction between the two paradigms lies in the nature of their
reproducibility guarantees. The traditional stack, anchored by Docker, provides
strong **spatial reproducibility**—the ability to execute the analysis
identically across different machines. However, it is vulnerable to **temporal
drift**, a concept that distinguishes reproducibility across space from
consistency over time [@malka2025space]. This temporal vulnerability arises from
several sources, including the resolution of mutable base image tags (e.g.,
`ubuntu:24.04`) and the potential for language-level package managers to resolve
dependencies differently over time.

The Nix-based approach is architected to solve this problem of temporal drift.
By pinning the entire software ecosystem to a single, immutable revision of the
`nixpkgs` repository, the framework provides strong guarantees of **temporal
reproducibility**. This claim is supported by large-scale empirical evidence; a
study by @malka2024yes, which rebuilt over 700,000 historical packages,
confirmed that Nix's functional model achieves a bit-for-bit reproducibility
rate exceeding 90% and a rebuildability rate over 99% across a multi-year
timeframe.

### Summary of Differences

| Dimension | Traditional Stack | Nix-Based Framework |
|-----------|------------------|---------------------|
| **Paradigm** | Imperative composition of specialized tools | Declarative specification in unified framework |
| **Environment setup** | Step-by-step shell commands | High-level specification |
| **Workflow definition** | File-based rules in Make syntax | Object-based graph in R syntax |
| **Scientific code** | Wrapped in CLI parsing and file I/O boilerplate | Pure functions with no plumbing |
| **File count** | 9 files (4 for orchestration and plumbing) | 6 files (2 for orchestration) |
| **Expertise required** | Linux system administration, Make syntax | R programming |
| **Spatial reproducibility** | Strong (via Docker isolation) | Strong (via Nix isolation) |
| **Temporal reproducibility** | Moderate (vulnerable to drift) | Strong (validated empirically [@malka2025]) |

This declarative rigor, however, is not without pragmatic costs. Our validation
on GitHub Actions shows that the imperative Docker-based runner completes the
entire pipeline in approximately three minutes, whereas the declarative
Nix-based runner takes around five minutes. This difference reflects the
trade-off between the upfront computational cost of Nix's rigorous dependency
resolution and the faster, but more brittle, imperative builds of the
traditional approach.

@@

The complete, runnable source code for this example is available in the paper's
GitHub repository. A shell script, `run_polyglot_example.sh`, is provided to
automate the entire process from environment creation to final output, as
described at the end of this
section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/repro-script/run_polyglot_example.sh}].
